---
title: "Galton example"
output: html_document
---

```{r load_hidden, echo=FALSE, results="hide", warning=FALSE}
suppressPackageStartupMessages({
  library(devtools)
  library(Biobase)
  library(UsingR)
})
```

```{r load}
  library(devtools)
  library(Biobase)
  library(UsingR)
```
# Galton-Example:Father Son Heights

To install these packages you can use the code (or if you are compiling the document, remove the `eval=FALSE` from the chunk.)

```{r install_packages, eval=FALSE}
install.packages(c("devtools","UsingR"))

```
The galton data is a set of data with father son heights. Let's plot it.

Load Galton Data

```{r galton,fig.height=4,fig.width=8}
data(galton)
par(mfrow=c(1,2))
hist(galton$child,col="blue",breaks=100)
hist(galton$parent,col="blue",breaks=100)
```

## The distribution of child heights

```{r, dependson="galton",fig.height=4,fig.width=4}
hist(galton$child,col="blue",breaks=100)
```

Let's look at the mean child's height  - average height:

```{r, dependson="galton",fig.height=4,fig.width=4}
hist(galton$child,col="blue",breaks=100)
meanChild <- mean(galton$child)
lines(rep(meanChild,100),seq(0,150,length=100),col="red",lwd=5)
```


What if we plot child versus average parent

```{r, dependson="galton",fig.height=4,fig.width=4}
plot(galton$parent,galton$child,pch=19,col="blue")
```

Average parent = 65 inches tall

```{r, dependson="galton",fig.height=4,fig.width=4}
plot(galton$parent,galton$child,pch=19,col="blue")
near65 <- galton[abs(galton$parent - 65)<1, ]
points(near65$parent,near65$child,pch=19,col="red")
lines(seq(64,66,length=100),rep(mean(near65$child),100),col="red",lwd=4)
```
## How would you predict the slope of the line for this data?

$$y = mx + b$$
