<<<<<<< HEAD
# Inference
=======
---
title: "Inference"
output: html_document
date: '2022-08-03'
---
>>>>>>> fa7b7a5b78cd4f7dfdc18fc75fa831b7c1427ae0

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<<<<<<< HEAD
=======
# Inference
>>>>>>> fa7b7a5b78cd4f7dfdc18fc75fa831b7c1427ae0

<a name="introduction"></a>


## Introduction: the process of reaching a conclusion from known facts. 

```{r,results=FALSE,echo=FALSE}
set.seed(1) #so that we get same results
```

The data introduced here is from a paper entitled The High-Fat Diet–Fed Mouse: A Model for Studying Mechanisms and Treatment of Impaired Glucose Tolerance and Type 2 Diabetes where they try and show that the mouse is a good model for studying diabetes and development of new treatments. the statistical concepts necessary to understand p-values and confidence intervals. These terms are ubiquitous in the life science literature. We'll use [this paper](https://doi.org/10.2337/diabetes.53.suppl_3.s215) as an example of . 

Note that the abstract has this statement: 

> "Body weight was higher in mice fed the high-fat diet already after the first week, due to higher dietary intake in combination with lower metabolic efficiency." 

To support this claim they provide the following in the results section:

> "Already during the first week after introduction of high-fat diet, body weight increased significantly more in the high-fat diet-fed mice ($+$ 1.6 $\pm$ 0.1 g) than in the normal diet-fed mice ($+$ 0.2 $\pm$ 0.1 g; P < 0.001)."

What does P < 0.001 mean? What are the $\pm$ included?
We will learn what this means and learn to compute these values in
R. The first step is to understand random variables. To do
this, we will use data from a mouse database (provided by Karen
Svenson via Gary Churchill and Dan Gatti and partially funded by P50
GM070683). We will import the data into R and explain random variables
and null distributions using R programming. 

If you already downloaded the `femaleMiceWeights` file into your working directory, you can read it into R with just one line:

```{r echo=FALSE, results="hide"}
library(downloader) ##use install.packages to install
dir <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/"
filename <- "femaleMiceWeights.csv"
url <- paste0(dir, filename)
if (!file.exists(filename)) download(url, destfile = filename)
```

```{r}
dat <- read.csv("femaleMiceWeights.csv")
```

Remember that a quick way to read the data, without downloading it is by using the url:

```{r,eval=FALSE}
dir <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/"
filename <- "femaleMiceWeights.csv"
url <- paste0(dir, filename)
dat <- read.csv(url)
```

** Our first look at data **

We are interested in determining if following a given diet makes mice
heavier after several weeks. This data was produced by ordering 24
mice from The Jackson Lab and randomly assigning either chow or high
fat (hf) diet. After several weeks, the scientists weighed each mouse
and obtained this data (`head` just shows us the first 6 rows):

```{r}
head(dat) 
``` 

In RStudio, you can view the entire dataset with:

```{r,eval=FALSE}
View(dat)
```

So are the hf mice heavier? Mouse 24 at 20.73 grams is one of the
lightest mice, while Mouse 21 at 34.02 grams is one of the heaviest. Both are on
the hf diet. Just from looking at the data, we see there is
*variability*. Claims such as the one above usually refer to the
averages. So let's look at the average of each group: 

```{r,message=FALSE}
library(dplyr)
control <- filter(dat,Diet == "chow") %>% select(Bodyweight) %>% unlist
treatment <- filter(dat,Diet == "hf") %>% select(Bodyweight) %>% unlist
print( mean(treatment) )
print( mean(control) )
obsdiff <- mean(treatment) - mean(control)
print(obsdiff)
```

So the hf diet mice are about 10% heavier. Are we done? Why do we need p-values and confidence intervals? The reason is that these averages are random variables. They can take many values. 

If we repeat the experiment, we obtain 24 new mice from The Jackson Laboratory and, after randomly assigning them to each diet, we get a different mean. Every time we repeat this experiment, we get a different value. We call this type of quantity a *random variable*. 

<a name="random_variable"></a>
## T.tests
** We can use a t.test to ask if there is a difference between the treatment and the control **
```{r}
t.test(treatment, control)
```

To see just the p-value, we can use the `$` extractor:

```{r}
result <- t.test(treatment,control)
result$p.value
```


