# Highthroughput Data

```{r, echo=FALSE}
library(knitr)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 40), tidy = TRUE)
```
## The data 

We want to move to inference when in comes to high-throughput data. High-throughput technologies have changed basic biology and the biomedical sciences from data poor disciplines to data intensive ones. A specific example comes from research fields interested in understanding gene expression. The Rmd file for this section is
available [here](https://github.com/gurinina/omic_sciences/blob/main/05-linear-modeling-genomics.Rmd). Gene expression is the process in which DNA, the blueprint for life, is copied into RNA, the templates for the synthesis of proteins, the building blocks for life. In the 1990s, the analysis of gene expression data amounted to spotting black dots on a piece of paper or extracting a few numbers from standard curves. With high-throughput technologies, such as microarrays, this suddenly changed to sifting through tens of thousands of numbers. More recently, RNA sequencing has further increased data complexity. Biologists went from using their eyes or simple summaries to categorize results, to having thousands (and now millions) of measurements per sample to analyze. In this chapter, we will focus on statistical inference in the context of high-throughput measurements. Specifically, we focus on the problem of detecting differences in groups using statistical tests and quantifying uncertainty in a meaningful way. We also introduce exploratory data analysis techniques that should be used in conjunction with inference when analyzing high-throughput data. In later chapters, we will study the statistics behind clustering, machine learning, factor analysis and multi-level modeling. 

Since there is a vast number of available public datasets, we use several gene expression examples. Nonetheless, the statistical techniques you will learn have also proven useful in other fields that make use of high-throughput technologies. Technologies such as microarrays, next generation sequencing, fMRI, and mass spectrometry all produce data to answer questions for which what we learn here will be indispensable. 

<a name="threetables"></a>

** Data packages**
Several of the  examples we are going to use in the following sections are best obtained through R packages. These are available from GitHub and can be installed using the `install_github` function from the `devtools` package. Microsoft Windows users might need to follow [these instructions](https://github.com/genomicsclass/windows) to properly install `devtools`. 

Once `devtools` is installed, you can then install the data packages like this:

```{r,eval=FALSE}
library(devtools)
library(GSE5859Subset)
install_github("genomicsclass/GSE5859Subset")
```


## The three tables in genomics

Most of the data we use as examples in this book are created with high-throughput technologies. These technologies measure thousands of _features_. Examples of features are genes, single base locations of the genome, genomic regions, or image pixel intensities. Each specific measurement product is defined by a specific set of features. For example, a specific gene expression microarray product is defined by the set of genes that it measures. 

A specific study will typically use one product to make measurements on several experimental units, such as individuals. The most common experimental unit will be the individual, but they can also be defined by other entities, for example different parts of a tumor. We often call the experimental units _samples_ following experimental jargon. It is important that these are not confused with samples as referred to in previous chapters, for example "random sample". 

So a high-throughput experiment is usually defined by three tables: one with the high-throughput measurements and two tables with information about the columns and rows of this first table respectively.

Because a dataset is typically defined by a set of experimental units and a product defines a fixed set of features, the high-throughput measurements can be stored in an $n \times m$ matrix, with $n$ the number of units and $m$ the number of features. In R, the convention has been to store the transpose of these matrices. 

Here is an example from a gene expression dataset:

```{r}
library(GSE5859Subset)
data(GSE5859Subset) ##this loads the three tables
dim(geneExpression)
```

We have RNA expression measurements for 8793 genes from blood taken from 24 individuals (the experimental units). For most statistical analyses, we will also need information about the individuals. For example, in this case the data was originally collected to compare gene expression across ethnic groups. However, we have created a subset of this dataset for illustration and separated the data into two groups:


```{r}
dim(sampleInfo)
head(sampleInfo)
sampleInfo$group
```

One of the columns, filenames, permits us to connect the rows of this table to the columns of the measurement table.

```{r}
match(sampleInfo$filename,colnames(geneExpression))
```


Finally, we have a table describing the features:

```{r}
dim(geneAnnotation)
head(geneAnnotation)
```

The table includes an ID that permits us to connect the rows of this table with the rows of the measurement table:
```{r}
head(match(geneAnnotation$PROBEID,rownames(geneExpression)))
```
The table also includes biological information about the features, namely chromosome location and the gene "name" used by biologists.

888888888888888888888888888888888888888888888888888888

## The Design Matrix

Here we will show how to use the two R functions, `formula`
and `model.matrix`, in order to produce *design matrices* (also known as *model matrices*) for a variety of linear models. We will use these design matrices when we model high dimenstional data. In the mouse diet examples we wrote the model as

$$ 
Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i, i=1,\dots,N 
$$

with $Y_i$ the weights 
and $x_i$ equal to 1 only when mouse $i$ receives the high fat diet. We use the term _experimental unit_ to $N$ different entities from which we obtain a measurement. In this case, the mice are the experimental units. 

This is the type of variable we will focus on in this chapter. We call them _indicator variables_ since they simply indicate if the experimental unit had a certain characteristic or not. As we described earlier, we can use linear algebra to represent this model:

$$
\mathbf{Y} = \begin{pmatrix}
Y_1\\
Y_2\\
\vdots\\
Y_N
\end{pmatrix}
,
\mathbf{X} = \begin{pmatrix}
1&x_1\\
1&x_2\\
\vdots\\
1&x_N
\end{pmatrix}
,
\boldsymbol{\beta} = \begin{pmatrix}
\beta_0\\
\beta_1
\end{pmatrix} \mbox{ and }
\boldsymbol{\varepsilon} = \begin{pmatrix}
\varepsilon_1\\
\varepsilon_2\\
\vdots\\
\varepsilon_N
\end{pmatrix}
$$



as: 


$$
\,
\begin{pmatrix}
Y_1\\
Y_2\\
\vdots\\
Y_N
\end{pmatrix} = 
\begin{pmatrix}
1&x_1\\
1&x_2\\
\vdots\\
1&x_N
\end{pmatrix}
\begin{pmatrix}
\beta_0\\
\beta_1
\end{pmatrix} +
\begin{pmatrix}
\varepsilon_1\\
\varepsilon_2\\
\vdots\\
\varepsilon_N
\end{pmatrix}
$$

or simply: 

$$
\mathbf{Y}=\mathbf{X}\boldsymbol{\beta}+\boldsymbol{\varepsilon}
$$

The design matrix is the matrix $\mathbf{X}$.

Once we define a design matrix, we are ready to find the least squares estimates. We refer to this as _fitting the model_. For fitting linear models in R, we will directly provide a _formula_ to the `lm` function. In this script, we will use the `model.matrix` function, which is used internally by the `lm` function. This will help us to connect the R `formula` with the matrix $\mathbf{X}$. It will therefore help us interpret the results from `lm`.

## Choice of design

The choice of design matrix is a critical step in linear modeling since it encodes which coefficients will be fit in the model, as well as the inter-relationship between the samples. 
A common misunderstanding is that the choice of design follows straightforward from a description of which samples were included in the experiment. This is not the case. The basic information about each sample (whether control or treatment group, experimental batch, etc.) does not imply a single 'correct' design matrix. The design matrix additionally encodes various assumptions about how the variables in $\mathbf{X}$ explain the observed values in $\mathbf{Y}$, on which the investigator must decide.

For the examples we cover here, we use linear models to make comparisons between different groups. Hence, the design matrices that we ultimately work with will have at least two columns: an _intercept_ column, which consists of a column of 1's, and a second column, which specifies which samples are in a second group. In this case, two coefficients are fit in the linear model: the intercept, which represents the population average of the first group, and a second coefficient, which represents the difference between the population averages of the second group and the first group. The latter is typically the coefficient we are interested in when we are performing statistical tests: we want to know if there is a difference between the two groups. Very often these come from the sampleInfo table in a summarized experiment.

We encode this experimental design in R with two pieces. We start with a formula with the tilde symbol `~`. This means that we want to model the observations using the variables to the right of the tilde. Then we put the name of a variable, which tells us which samples are in which group.

Let's try an example. Suppose we have two groups, control and high fat diet, with two samples each. For illustrative purposes, we will code these with 1 and 2 respectively. We should first tell R that these values should not be interpreted numerically, but as different levels of a *factor*. We can then use the paradigm `~ group` to, say, model on the variable `group`.

```{r}
group <- factor( c(1,1,2,2) )
model.matrix(~ group)
```

(Don't worry about the `attr` lines printed beneath the matrix. We won't be using this information.)

What about the `formula` function? We don't have to include this. By starting an expression with `~`, it is equivalent to telling R that the expression is a formula:

```{r}
model.matrix(formula(~ group))
```

What happens if we don't tell R that `group` should be interpreted as a factor?

```{r}
group <- c(1,1,2,2)
model.matrix(~ group)
```

This is **not** the design matrix we wanted, and the reason is that we provided a numeric variable as opposed to an _indicator_ to the `formula` and `model.matrix` functions, without saying that these numbers actually referred to different groups. We want the second column to have only 0 and 1, indicating group membership.

A note about factors: the names of the levels are irrelevant to `model.matrix` and `lm`. All that matters is the order. For example:

```{r}
group <- factor(c("control","control","highfat","highfat"))
model.matrix(~ group)
```

produces the same design matrix as our first code chunk.

### More groups

Using the same formula, we can accommodate modeling more groups. Suppose we have a third diet:

```{r}
group <- factor(c(1,1,2,2,3,3))
model.matrix(~ group)
```

Now we have a third column which specifies which samples belong to the third group.

An alternate formulation of design matrix is possible by specifying `+ 0` in the formula:

```{r}
group <- factor(c(1,1,2,2,3,3))
model.matrix(~ group + 0)
```

This group now fits a separate coefficient for each group. We will explore this design in more depth later on.

** More variables **

We have been using a simple case with just one variable (diet) as an example. In the life sciences, it is quite common to perform experiments with more than one variable. For example, we may be interested in the effect of diet and the difference in sexes. In this case, we have four possible groups:

```{r}
diet <- factor(c(1,1,1,1,2,2,2,2))
sex <- factor(c("f","f","m","m","f","f","m","m"))
table(diet,sex)
```

If we assume that the diet effect is the same for males and females (this is an assumption), then our linear model is:

$$
Y_{i}= \beta_0 + \beta_1 x_{i,1} + \beta_2 x_{i,2} + \varepsilon_i 
$$

To fit this model in R, we can simply add the additional variable with a `+` sign in order to build a design matrix which fits based on the information in additional variables:

```{r}
diet <- factor(c(1,1,1,1,2,2,2,2))
sex <- factor(c("f","f","m","m","f","f","m","m"))
model.matrix(~ diet + sex)
```

The design matrix includes an intercept, a term for `diet` and a term for `sex`. We would say that this linear model accounts for differences in both the group and condition variables. However, as mentioned above, the model assumes that the diet effect is the same for both males and females. We say these are an _additive_ effect. For each variable, we add an effect regardless of what the other is. Another model is possible here, which fits an additional term and which encodes the potential interaction of group and condition variables. We will cover interaction terms in depth in a later script.

The interaction model can be written in either of the following two formulas:

```{r,eval=FALSE}
model.matrix(~ diet + sex + diet:sex)
```

or 

```{r}
model.matrix(~ diet*sex)
```

** Releveling **

The level which is chosen for the *reference level* is the level which is contrasted against.  By default, this is simply the first level alphabetically. We can specify that we want group 2 to be the reference level by either using the `relevel` function:

```{r}
group <- factor(c(1,1,2,2))
group <- relevel(group, "2")
model.matrix(~ group)
```

or by providing the levels explicitly in the `factor` call:
```{r}
group <- factor(group, levels = c("1","2"))
model.matrix(~ group)
```
888888888888888888888888888888888888888888888888888888

## Linear Regression in practice
Linear regression is used to predict the value of an outcome variable Y based on one or more input predictor variables X. The aim is to establish a linear relationship (a mathematical formula) between the predictor variable(s) and the response variable, so that, we can use this formula to estimate the value of the response Y, when only the predictors (Xs) values are known.

Introduction
The aim of linear regression is to model a continuous variable Y as a mathematical function of one or more X variable(s), so that we can use this regression model to predict the Y when only the X is known. This mathematical equation can be generalized as follows:

$$ \mathbf{Y} = {\beta1} + {\beta2}\mathbf{X} + {\epsilon} $$


where, β1 is the intercept and β2 is the slope. Collectively, they are called regression coefficients. ϵ is the error term, the part of Y the regression model is unable to explain.

Example Problem
For this analysis, we will use the cars dataset that comes with R by default. cars is a standard built-in dataset, that makes it convenient to demonstrate linear regression in a simple and easy to understand fashion. You can access this dataset simply by typing in cars in your R console. You will find that it consists of 50 observations(rows) and 2 variables (columns) – dist and speed. Lets print out the first six observations here..

```{r}
head(cars)  
```

Before we begin building the regression model, it is a good practice to analyze and understand the variables. The graphical analysis and correlation study below will help with this.

Graphical Analysis
The aim of this exercise is to build a simple regression model that we can use to predict Distance (dist) by establishing a statistically significant linear relationship with Speed (speed). But before jumping in to the syntax, lets try to understand these variables graphically. Typically, for each of the independent variables (predictors), the following plots are drawn to visualize the following behavior:

Scatter plot: Visualize the linear relationship between the predictor and response

Box plot: To spot any outlier observations in the variable. Having outliers in your predictor can drastically affect the predictions as they can easily affect the direction/slope of the line of best fit.
Density plot: To see the distribution of the predictor variable. Ideally, a close to normal distribution (a bell shaped curve), without being skewed to the left or right is preferred. Let us see how to make each one of them.

Scatter Plot
Scatter plots can help visualize any linear relationships between the dependent (response) variable and independent (predictor) variables. Ideally, if you are having multiple predictor variables, a scatter plot is drawn for each one of them against the response, along with the line of best as seen below.

```{r}
scatter.smooth(x = cars$speed, y = cars$dist, main = "Dist ~ Speed")  # scatterplot
```

The scatter plot along with the smoothing line above suggests a linearly increasing relationship between the ‘dist’ and ‘speed’ variables. This is a good thing, because, one of the underlying assumptions in linear regression is that the relationship between the response and predictor variables is linear and additive.

BoxPlot – Check for outliers
Generally, any datapoint that lies outside the 1.5 * interquartile-range (1.5 * IQR) is considered an outlier, where, IQR is calculated as the distance between the 25th percentile and 75th percentile values for that variable.

```{r}
par(mfrow = c(1, 2))  # divide graph area in 2 columns
boxplot(cars$speed, main = "Speed", sub = paste("Outlier rows: ", boxplot.stats(cars$speed)$out))  # box plot for 'speed'
boxplot(cars$dist, main = "Distance", sub = paste("Outlier rows: ", boxplot.stats(cars$dist)$out))  # box plot for 'distance'
```
Check if the response variable is close to normality
Normal Q-Q plots take your sample data, sort it in ascending order, and then plot them versus quantiles calculated from a theoretical normal distribution. The number of quantiles is selected to match the size of your sample data. In R this can be done using the `qnorm` and `qline` functions. `qnorm` plots your data agaist the theoretical normal distribution, and `qline` plots the normal distribution against itself for comparison.  

```{r}
library(rafalib)
mypar(1,2) # divide graph area in 2 columns
par(pch = 19)  
qqnorm(cars$speed, ylab = "speed quantiles",xlab = "normal quantiles",
       main = "Speed Q-Q Plot", col = "dodgerblue")
qqline(cars$speed)
qqnorm(cars$dist, ylab = "distance quantiles",xlab = "normal quantiles",
       main = "Distance Q-Q Plot",col = "dodgerblue")
qqline(cars$dist)

```

Correlation
Correlation is a statistical measure that suggests the level of linear dependence between two variables, that occur in pair – just like what we have here in speed and dist. Correlation can take values between -1 to +1. If we observe for every instance where speed increases, the distance also increases along with it, then there is a high positive correlation between them and therefore the correlation between them will be closer to 1. The opposite is true for an inverse relationship, in which case, the correlation between the variables will be close to -1.

A value closer to 0 suggests a weak relationship between the variables. A low correlation (-0.2 < x < 0.2) probably suggests that much of variation of the response variable (Y) is unexplained by the predictor (X), in which case, we should probably look for better explanatory variables.

```{r}
cor(cars$speed, cars$dist)  # calculate correlation between speed and distance 
```

Build Linear Model
Now that we have seen the linear relationship pictorially in the scatter plot and by computing the correlation, lets see the syntax for building the linear model. The function used for building linear models is lm(). The lm() function takes in two main arguments, namely: 1. Formula 2. Data. The data is typically a data.frame and the formula is a object of class formula. But the most common convention is to write out the formula directly in place of the argument as written below.

```{r}
linearMod <- lm(dist ~ speed, data = cars)  # build linear regression model on full data
print(linearMod)
```

Now that we have built the linear model, we also have established the relationship between the predictor and response in the form of a mathematical formula for Distance (dist) as a function for speed. For the above output, you can notice the ‘Coefficients’ part having two components: Intercept: -17.579, speed: 3.932 These are also called the beta coefficients. In other words,

$$ dist = Intercept + ({\beta} * speed) => dist = −17.579 + 3.932∗speed $$



Linear Regression Diagnostics

Now the linear model is built and we have a formula that we can use to predict the dist value if a corresponding speed is known. Is this enough to actually use this model? NO! Before using a regression model, you have to ensure that it is statistically significant. How do you ensure this? Lets begin by printing the summary statistics for linearMod.

```{r}
summary(linearMod)  # model summary
```

The p Value: Checking for statistical significance
The summary statistics above tells us a number of things. One of them is the model p-Value (bottom last line) and the p-Value of individual predictor variables (extreme right column under ‘Coefficients’). The p-Values are very important because, We can consider a linear model to be statistically significant only when both these p-Values are less that the pre-determined statistical significance level, which is ideally 0.05. This is visually interpreted by the significance stars at the end of the row. The more the stars beside the variable’s p-Value, the more significant the variable.

Null and alternate hypothesis

When there is a p-value, there is a hull and alternative hypothesis associated with it. In Linear Regression, the Null Hypothesis is that the coefficients associated with the variables is equal to zero. The alternate hypothesis is that the coefficients are not equal to zero (i.e. there exists a relationship between the independent variable in question and the dependent variable).

t-value
We can interpret the t-value something like this. A larger t-value indicates that it is less likely that the coefficient is not equal to zero purely by chance. So, higher the t-value, the better.

Pr(>|t|) or p-value is the probability that you get a t-value as high or higher than the observed value when the Null Hypothesis (the β coefficient is equal to zero or that there is no relationship) is true. So if the Pr(>|t|) is low, the coefficients are significant (significantly different from zero). If the Pr(>|t|) is high, the coefficients are not significant.


What this means to us? when p Value is less than significance level (< 0.05), we can safely reject the null hypothesis that the co-efficient β ${\beta}$ of the predictor is zero. In our case, linearMod, both these p-Values are well below the 0.05 threshold, so we can conclude our model is indeed statistically significant.

It is absolutely important for the model to be statistically significant before we can go ahead and use it to predict (or estimate) the dependent variable, otherwise, the confidence in predicted values from that model reduces and may be construed as an event of chance.

How to calculate the t Statistic and p-Values?

When the model co-efficients and standard error are known, the formula for calculating t Statistic and p-Value is as follows:

$$t−Statistic = \frac{{\beta}−coefficient}{Std.Error} $$


```{r}
modelSummary <- summary(linearMod)  # capture model summary as an object
modelCoeffs <- modelSummary$coefficients  # model coefficients
beta.estimate <- modelCoeffs["speed", "Estimate"]  # get beta estimate for speed
std.error <- modelCoeffs["speed", "Std. Error"]  # get std.error for speed
t_value <- beta.estimate/std.error  # calc t statistic
p_value <- 2*pt(-abs(t_value), df = nrow(cars)-ncol(cars))  # calc p Value
f_statistic <- linearMod$fstatistic[1]  # fstatistic
f <- summary(linearMod)$fstatistic  # parameters for model p-value calc
model_p <- pf(f[1], f[2], f[3], lower.tail = FALSE)
```

R-Squared 
The actual information in a data is the total variation it contains, remember?. What R-Squared tells us is the proportion of variation in the dependent (response) variable that has been explained by this model. In our example, 
the $R^2$ we get is 0.6510794. Or roughly 65% of the variance found in the response variable (dist) can be explained by the predictor variable (speed). Pretty good, so speed is a good predictor of the distance required for a car to stop. A reminder of where $R^2$ comes from:


$$ R^2 = 1−\frac{SSE}{SST}$$
where SSE is the sum of squared errors given by:
$$SSE= \sum_{i = 1}^{n}{(y_i - \hat{y_i})^2}$$
and SST is the sum of squared total:
$$SST = \sum_{i = 1}^{n}{(y_i - \bar{y_i})^2}$$

Here, $\hat{y_i}$ is the fitted value for observation i and $\bar{y_i}$ is the mean of Y.

Finally, we can plot the results:

```{r}
par(pch = 19)
plot(cars$speed,cars$dist,ylab = "distance",xlab = "speed",
       main = "Stopping Distance as a Function of Speed")
points(cars$speed,linearMod$fitted.values,col = "dodgerblue")
lines(cars$speed,linearMod$fitted.values,col = "dodgerblue")
```


888888888888888888888888888888888888888888888888888888
## The mouse diet example

We will demonstrate how to analyze the high fat diet data using linear models instead of directly applying a t-test. We will demonstrate how ultimately these two approaches are equivalent. 

We start by reading in the data and creating a quick stripchart:

```{r,echo=FALSE}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleMiceWeights.csv"
filename <- "femaleMiceWeights.csv"
library(downloader)
if (!file.exists(filename)) download(url, filename)
```

```{r,echo=FALSE}
set.seed(1) #same jitter in stripchart
```

```{r bodyweight_by_diet_stripchart, fig.cap="Mice bodyweights stratified by diet."}
dat <- read.csv("femaleMiceWeights.csv") ## previously downloaded
par(pch = 22)
stripchart(dat$Bodyweight ~ dat$Diet, vertical=TRUE, method="jitter",
           main="Bodyweight over Diet",pch = 19)
```

We can see that the high fat diet group appears to have higher weights on average, although there is overlap between the two samples.

For demonstration purposes, we will build the design matrix $\mathbf{X}$ using the formula `~ Diet`. The group with the 1's in the second column is determined by the level of `Diet` which comes second; that is, the non-reference level. 

```{r}
levels(dat$Diet)
X <- model.matrix(~ Diet, data=dat)
head(X)
```

## The Mathematics Behind lm()

Before we use our shortcut for running linear models, `lm`, we want to review what will happen internally. Inside of `lm`, we will form the design matrix $\mathbf{X}$ to solve the linear equation:

$$ \mathbf{Y} = \hat{\boldsymbol{\beta}}\mathbf{X} $$

and calculate the $\boldsymbol{\beta}$, which minimizes the sum of squares using the previously described formula. The formula for this solution is:

$$ \hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \mathbf{X})^{-1} \mathbf{X}^\top \mathbf{Y} $$

We can calculate this in R using our matrix multiplication operator `%*%`, the inverse function `solve`, and the transpose function `t`.


```{r}
Y <- dat$Bodyweight
X <- model.matrix(~ Diet, data=dat)
solve(t(X) %*% X) %*% t(X) %*% Y

solveLM = function(Y, design){
        
    sv =   solve(t(design) %*% design) %*% t(design) %*% Y  
    sv
}
# just to go back to the cars example for a second...model matrices aren't really necesssary when you have just two variables and all your variable are numeric (covariates) as opposed being a factor
d = model.matrix(dist~speed,cars)
solveLM(Y=cars$dist,design = d)
```
These coefficients are the average of the control group and the difference of the averages:


```{r}
s <- split(dat$Bodyweight, dat$Diet)
mean(s[["chow"]])
mean(s[["hf"]]) - mean(s[["chow"]])
```

Finally, we use our shortcut, `lm`, to run the linear model:

```{r}
fit <- lm(Bodyweight ~ Diet, data=dat)
summary(fit)
(coefs <- coef(fit))

# We can use the model matrix without the intercept to do the same as lm includes an offset in the function:

fitY<- lm(Y ~ 0 + X)
summary(fitY)$coefficients

# For genomics data, where y can be a matrix and where each row is a gene and each column is an experimental sample. Each y is then independently fitting a gene in the rows of the matrix.
fit.Y <-  lm.fit(x = X,y = Y)
fit.Y$coefficients
```
**Examining the coefficients**

The following plot provides a visualization of the meaning of the coefficients with colored arrows (code not shown):

```{r parameter_estimate_illustration, fig.cap="Estimated linear model coefficients for bodyweight data illustrated with arrows.",echo=FALSE}
stripchart(dat$Bodyweight ~ dat$Diet, vertical=TRUE, method="jitter",
           main="Bodyweight over Diet", ylim=c(0,40), xlim=c(0,3),pch = 19)
a <- -0.25
lgth <- .1
library(RColorBrewer)
cols <- brewer.pal(3,"Dark2")
abline(h=0)
arrows(1+a,0,1+a,coefs[1],lwd=3,col=cols[1],length=lgth)
abline(h=coefs[1],col=cols[1])
arrows(2+a,coefs[1],2+a,coefs[1]+coefs[2],lwd=3,col=cols[2],length=lgth)
abline(h=coefs[1]+coefs[2],col=cols[2])
legend("right",names(coefs),fill=cols,cex=.75,bg="white")
```

To make a connection with material presented earlier, this simple linear model is actually giving us the same result (the t-statistic and p-value) for the difference as a specific kind of t-test. This is the t-test between two groups with the assumption that the population standard deviation is the same for both groups. This was encoded into our linear model when we assumed that the errors $\boldsymbol{\varepsilon}$ were all equally distributed.

Although in this case the linear model is equivalent to a t-test, we will soon explore more complicated designs, where the linear model is a useful extension. Below we demonstrate that one does in fact get the exact same results:

Our `lm` estimates were:

```{r}
summary(fit)$coefficients
```

**And the t-statistic  is the same:**

```{r}
ttest <- t.test(s[["hf"]], s[["chow"]], var.equal=TRUE)
summary(fit)$coefficients[2,3]
ttest$statistic
```

888888888888888888888888888888888888888888888888888888

## Multiple comparisons 

Most scientifically compelling questions concern biological variation,
which can only be directly measured with good experimental design.  Accurate
interpretation of origin and size of biological variation requires
appropriate statistical analysis.

In this section we will cover inference in the context of genome-scale experiments.  There are several serious conceptual problems:

- there are many tests, often at least one test for each one of tens of thousands of features
- each feature (typically a gene) exhibits its own technical and biological variability
- there may be unmeasured or unreported sources of biological variation (such as time of day)
- many features are inherently interrelated, so the tests are not independent

888888888888888888888888888888888888888888888888888888


## Inference in Practice

Suppose we were given high-throughput gene expression data that was measured for several individuals in two populations. We are asked to report which genes have different average expression levels in the two populations. If instead of thousands of genes, we were handed data from just one gene, we could simply apply the inference techniques that we have learned before. We could, for example, use a t-test or some other test. Here we review what changes when we consider high-throughput data.

### p-values are random variables

An important concept to remember in order to understand the concepts presented in this chapter is that p-values are random variables. To see this,  consider the example in which we define a p-value from a t-test with a large enough sample size to use the CLT approximation. Then our p-value is defined as the probability that a normally distributed random variable is larger, in absolute value, than the observed t-test, call it $Z$. So for a two sided test the p-value is: 

$$
p = 2 \{ 1 - \Phi(\mid Z \mid)\}
$$

In R, we write:
```{r,eval=FALSE}
2*( 1-pnorm( abs(Z) ) )
```

Now because $Z$ is a random variable and $\Phi$ is a deterministic
function, $p$ is also a random variable. We will create a Monte Carlo
simulation showing how the values of $p$ change. We use `femaleControlsPopulation.csv` from earlier chapters.

```{r,echo=FALSE}
filename <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleControlsPopulation.csv"
```

We read in the data, and use `replicate` to repeatedly create p-values.

```{r pvalue_hist, fig.cap="P-value histogram for 10,000 tests in which null hypothesis is true."}
set.seed(1)
population = unlist( read.csv(filename) )
N <- 12
B <- 10000
pvals <- replicate(B,{
  control = sample(population,N)
  treatment = sample(population,N)
  t.test(treatment,control)$p.val 
  })
hist(pvals)
```

As implied by the histogram, in this case the distribution of the p-value is uniformly distributed. In fact, we can show theoretically that when the null hypothesis is true, this is always the case. For the case in which we use the CLT, we have that the null hypothesis $H_0$ implies that our test statistic $Z$  follows a normal distribution with mean 0 and SD 1 thus:

$$
p_a = \mbox{Pr}(Z < a \mid H_0) = \Phi(a)
$$

This implies that:

$$
\begin{aligned}
\mbox{Pr}(p < p_a) &= \mbox{Pr}[ \Phi^{-1}(p) < \Phi^{-1}(p_a) ] \\
  & = \mbox{Pr}(Z < a) = p_a
\end{aligned}
$$

which is the definition of a uniform distribution.

## Thousands of tests

In this data we have two groups denoted with 0 and 1:
```{r}
library(GSE5859Subset)
data(GSE5859Subset)
g <- sampleInfo$group
g
```

If we were interested in a particular gene, let's arbitrarily pick the one on the 25th row, we would simply compute a t-test. To compute a p-value, we will use the t-distribution approximation and therefore we need the population data to be approximately normal. We check this assumption with a qq-plot:

```{r qqplots_for_one_gene, fig.cap="Normal qq-plots for one gene. Left plot shows first group and right plot shows second group.",fig.width=10.5,fig.height=5.25}
e <- geneExpression[25,]
library(rafalib)
mypar(1,2)
qqnorm(e[g==1])
qqline(e[g==1])
qqnorm(e[g==0])
qqline(e[g==0])
```

The qq-plots show that the data is well approximated by the normal approximation. The t-test does not find this gene to be statistically significant:

```{r}
t.test(e[g==1],e[g==0])$p.value
```

To answer the question for each gene, we simply repeat the above for each gene. Here we will define our own function and use `apply`:

```{r}
myttest <- function(x) t.test(x[g==1],x[g==0],var.equal=TRUE)$p.value
pvals <- apply(geneExpression,1,myttest)
```

We can now see which genes have p-values less than, say, 0.05. For example, right away we see that...

```{r}
sum(pvals<0.05)
```

... genes had p-values less than 0.05.

However, as we will describe in more detail below, we have to be careful in interpreting this result because we have performed over 8,000 tests. If we performed the same procedure on random data, for which the null hypothesis is true for all features, we obtain the following results:

```{r}
set.seed(1)
m <- nrow(geneExpression)
n <- ncol(geneExpression)
randomData <- matrix(rnorm(n*m),m,n)
nullpvals <- apply(randomData,1,myttest)
sum(nullpvals<0.05)
```

As we will explain later in the chapter, this is to be expected: 419 is roughly 0.05*8192 and we will describe the theory that tells us why this prediction works.

### Faster t-test implementation

Before we continue, we should point out that the above implementation is very inefficient. There are several faster implementations that perform t-test for high-throughput data. We make use of a function that is not available from CRAN, but rather from the Bioconductor project. 

To download and install packages from Bioconductor, we can use the `install_bioc` function in `rafalib` to install the package:


```{r,eval=FALSE}
install_bioc("genefilter")
```

Now we can show that this function is much faster than our code above and produce practically the same answer:

```{r,message=FALSE}
library(genefilter)
results <- rowttests(geneExpression,factor(g))
max(abs(pvals-results$p))
```

